---
title: "Data Tables and Quality Control"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Data Tables and Quality Control}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r chunk, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.width = 7,
    fig.height = 5,
    out.width = "100%",
    dev = "ragg_png"
)
```

```{r setup}
library(mesa)
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
```

# Generating tables of values from a qseaSet
`qsea` has several different types of values that can be returned per window.
You can get a table of raw counts per window with `getCountTable`:

```{r generating-tables-of-values-fr}
exampleTumourNormal %>%
    getCountTable()
```

but this is not particularly useful, as different samples have generally been
sequenced to different read depths. Note that the annotation of the `qseaSet`
regions is included in this output, notably the `CpG_density` that is
described in `vignette("generation", package = "mesa")`.

The next option is normalised reads per million (NRPM). This normalises the
number of reads by the depth of sequencing, as well as by copy number variation.

```{r chunk-2}
exampleTumourNormal %>%
    getNRPMTable()
```

Finally, we can request beta values. Beta values are measures of methylation,
scaled to lie between 0 and 1, where 0 means that the window is unmethylated and
1 is methylated. These are scaled with respect to a number of further metrics,
including the region's `CpG_density`, background noise and the sample enrichment
profile, and are therefore the most useful value to consider when comparing
methylation between samples.

```{r chunk-3}
exampleTumourNormal %>%
    getBetaTable()
```

Note that for some windows an NA value will be returned. This occurs when fewer
than three reads would be required for the window to be fully methylated, at the
read depth and enrichment profile calculated per sample. These occur
particularly at windows with a low `CpG_density`.

All three of these functions take an option `useGroupMeans` that determines
whether to average over the `group` column of the qseaSet `sampleTable`, which
can be useful when you have biological or technical replicates for example.

In the example qseaSet, `group` is the same as `sample_name`, so we need to
change that to see the effect:

```{r chunk-4}
exampleTumourNormal %>%
    getBetaTable(useGroupMeans = FALSE)

exampleTumourNormal %>%
    mutate(group = str_remove(sample_name,"\\d")) %>%
    getBetaTable(useGroupMeans = TRUE) 
```

Now the table has one column per group (four here) rather than one per sample.
Windows where no sample in a group has a beta value are returned as `NaN`.

We explicitly note that qsea incorporates the uncertainty of the measurements in
its beta value estimations. This means that the values returned as `beta` values
are actually the median of this estimate, with an associated confidence interval
(which can be queried, see `qsea::makeTable`). In practice, this means that
samples with high coverage can return values very close to 0 and 1, but the same
sample with less coverage might return 0.1 or 0.9. This can have an effect on
clustering and PCAs for example.

# Quality Control
One extremely important aspect of any sequencing experiment is quality control
(QC), and ME-seq is no exception. Alongside the typical QC steps on the bam
files themselves (although a raised CG content should obviously be expected
following enrichment), there are several methylation-specific steps to consider.

## Relative Enrichment 
The first metric that we use is a relative enrichment metric, known as `relH`.
This was first introduced in the `MEDIPS` package, via the `MEDIPS.CpGenrich()`
function, and is a metric to express the proportion of CG dinucleotides,
compared to the reference genome. The relative enrichment `relH` is calculated
in the following way:

* For each mapped DNA fragment, retrieve the end-to-end DNA spanned on the
    reference sequence. This is used instead of the reads themselves, as a
    fragment will often be longer than the read(s) sequenced at its ends, and
    may therefore contain CGs not covered by the reads.
* In these sequences, count how many CGs are present in total, and divide by the
    total number of base pairs covered.
* Divide by the same ratio, calculated for the entire genome.

This is therefore a measure of how many more CGs are present in the fragments
covered within the bam file, compared to the reference genome.

Within `mesa`, this metric is calculated through `makeQset`, as part of the
information added to the `libraries` table on the `qseaSet`. For the MBD-seq
data that we have studied, we typically use a minimum `relH` of 2.5, with 3 as
a stricter cutoff. This value is typically around 4-5 in many of our studies.

We can use the `addLibraryInformation` function to add the information in the
`libraries` table directly onto the `sampleTable`:

```{r chunk-5}
exampleTumourNormal %>% 
    addLibraryInformation() %>% 
    getSampleTable() %>% 
    dplyr::select(sample_name, relH)
```

When a non-enriched ("Input") sample is used to calculate copy number variation,
an equivalent calculation is also performed on these non-enriched fragments.
This is also stored in the `libraries` table, in columns starting with
`input_`:

```{r chunk-6}
exampleTumourNormal %>% 
    addLibraryInformation() %>% 
    getSampleTable() %>% 
    dplyr::select(sample_name, input_relH)
```

We typically see values of `input_relH` between 1.3 and 1.6, suggesting that
there are more CGs seen in the sequenced fragments than would be expected from
the genomic sequence. This is possibly due to the evolutionary suppression of
`CG` dinucleotides outside of specific contexts, meaning they are less common in
those areas of the genome that are repetitive and hence less easy to map. This
is likely to depend on the particular sequencing methods and the sample
type under consideration. We therefore need to see an increase between the
enriched and non-enriched data:

```{r chunk-7}
exampleTumourNormal %>% 
    addLibraryInformation() %>% 
    getSampleTable() %>% 
    dplyr::select(sample_name, relH_MeCap = relH, relH_Input = input_relH) %>% 
    pivot_longer(matches("relH"), 
                names_prefix = "relH_", 
                names_to = "enrichment", 
                values_to = "relH") %>%
    ggplot(aes(x = sample_name, y = relH, col = enrichment)) +
    geom_point() +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

There is a second relative enrichment metric defined by the `MEDIPS` package,
known as `GoGe`, that is also calculated globally in a similar way but considers
the ratio between the number of `CG`s present in the fragments compared to the
number of `C`s and `G`s independently. This is calculated and stored in the
`libraries` information alongside the `relH` metric.

## Fragments without pattern
The `libraries` table also records how many fragments do not contain a CG at
all. This can be a measure of background noise.

## Hyper-stably methylated regions
The other metric that we typically use for quality control is to consider
windows that are known to be commonly methylated across a range of tissues and
diseases. For instance, for the human genome, [Edgar et al.
(2014)](https://doi.org/10.1186/1756-8935-7-28) identified probes from the
Illumina 450k methylation array that were consistently methylated in over 1000
samples. We therefore take the windows covered by these probes, calculate
beta values for them, and then determine how many of these windows have a
calculated beta value above a threshold (e.g. 0.8). As noted above, beta values
will not be calculated if a sample does not have sufficient coverage and/or
enrichment, meaning this metric incorporates both the coverage depth as well as
how well the enrichment has worked. For our samples within the National
Biomarker Centre, we typically look for at least 40% of the hyper-stable
windows with a `CpG_density` above 5 to have a beta value above 0.8. Samples
with lower enrichment scores (as described by the `relH` score) require a
greater coverage depth to reach this threshold.

For hg38, these hyper-stable methylated probes are provided in a data object,
`hg38UltraStableProbes`, and the function `addHyperStableFraction` calculates
this fraction and adds it as a column, `hyperStableEdgar`, to the sampleTable
for filtering. Note that this will not work with the provided example data, as
there is no overlap with the windows in the example dataset (i.e.
`filterByOverlaps(exampleTumourNormal, hg38UltraStableProbes)` has no windows).

## Summary
We find that these three metrics, `valid_fragments`, `relH` and
`hyperStableEdgar` are the most useful for quality control. As a rough guide,
for 300bp windows on the human genome with T7-MBD-seq, we typically look for a
`relH` of at least 2.5, a `hyperStableEdgar` value of at least 0.4 (40%) and
around 3,000,000 `valid_fragments`, although most samples that pass the first
two will reach this anyway. We apply these as guidelines rather than strict
cutoffs, and suitable values will depend on the enrichment protocol (MBD-seq or
MeDIP-seq), the sample type and the sequencing depth, so we recommend choosing
thresholds from the distribution of these metrics in your own data.


## Session info
```{r session-info, echo=FALSE}
sessionInfo()
```
